---
title: "Working with Spatial Data in R"
subtitle: BIOL 806- assignment 8
date: 04/30/2026
author: Lala Movsesian
format: html
---

# Introduction

These are example data showing regions, rivers, and population in Alaska.
Data source: https://dev.nceas.ucsb.edu/view/urn:uuid:6f07cb25-a4a1-48e8-95cb-74f532f3ce2d

# Setup

Loading libraries:

```{r}
library(readr)
library(here)
library(sf)
library(scales)
library(dplyr)
library(tidyr)
library(ggplot2)
library(leaflet) 
library(ggspatial)
library(prettymapr)
```
loading data:
```{r}
ak_rgns_sf <- read_sf(here("data/raw/ak_regions_simp.shp"))
rivers_3338_sf <- read_sf(here("data/raw/ak_rivers_simp.shp"))
```


# Exploring data 




transforming:

```{r}
ak_rgns_3338_sf <- ak_rgns_sf %>%
  st_transform(crs = 3338)
st_crs(ak_rgns_3338_sf)
```
plot:

```{r}
plot(ak_rgns_3338_sf)
```




population data:

```{r}
pop_df <- read_csv(here("data/raw/alaska_population.csv"))
```

prep for join (transform to sf)

```{r}
pop_4326_sf <- st_as_sf(pop_df,
                        coords = c('lng', 'lat'),
                        crs = 4326,
                        remove = F)

```

```{r}
pop_3338_sf <- st_transform(pop_4326_sf, 
                            crs = 3338)
```



joining population data with region data:

```{r}
pop_joined_sf <- st_join(pop_3338_sf, 
                         ak_rgns_3338_sf, 
                         join = st_within)
```


calculate population:

```{r}
pop_rgn_df <- pop_joined_sf %>%
    st_drop_geometry() %>%
    group_by(region) %>%
    summarize(total_pop = sum(population))
```



```{r}
pop_rgn_3338_sf <- left_join(ak_rgns_3338_sf, 
                             pop_rgn_df, 
                             by = "region")

plot(pop_rgn_3338_sf["total_pop"])
```


```{r}
pop_mgmt_3338_sf <- pop_rgn_3338_sf %>%
    group_by(mgmt_area) %>%
    summarize(total_pop = sum(total_pop))

plot(pop_mgmt_3338_sf["total_pop"])
```

```{r}
pop_mgmt_3338_sf <- pop_rgn_3338_sf %>%
    group_by(mgmt_area) %>%
    summarize(total_pop = sum(total_pop), do_union = F)

plot(pop_mgmt_3338_sf["total_pop"])
```



saving the file:
```{r}
write_sf(pop_rgn_3338_sf, here("data/ak_regions_population.shp"))
```

# ggplot

```{r}
ggplot(pop_rgn_3338_sf) +
    geom_sf(aes(fill = total_pop)) +
    labs(fill = "Total Population") +
    scale_fill_continuous(low = "khaki",
                          high =  "firebrick",
                          labels = comma) +
    theme_bw()
```



checking EPSG code in river data:
```{r}
st_crs(rivers_3338_sf)
```

plotting population with rivers


```{r}
ggplot() +
    geom_sf(data = pop_rgn_3338_sf, 
            aes(fill = total_pop)) +
    geom_sf(data = pop_3338_sf, 
            size = 0.5) +
    geom_sf(data = rivers_3338_sf,
            aes(linewidth = StrOrder)) +
    scale_linewidth(range = c(0.05, 0.5),
                    guide = "none") +
    labs(title = "Total Population by Alaska Region",
         fill = "Total Population") +
    scale_fill_continuous(low = "khaki",
                          high =  "firebrick",
                          labels = comma) +
    theme_bw() 
```

transforming ESPG code to add base map;
```{r}
pop_3857_sf <- st_transform(pop_3338_sf, 
                         crs = 3857)
```

adding base map:

```{r}
ggplot(data = pop_3857_sf) +
    ggspatial::annotation_map_tile(type = "osm", zoom = 4, progress = 'none') + 
    geom_sf(aes(color = population),
            fill = NA) +
    scale_color_continuous(low = "darkkhaki",
                           high =  "firebrick",
                           labels = comma)
```



# Leaflet

defining a projection:

```{r}
epsg3338 <- leaflet::leafletCRS(
    crsClass = "L.Proj.CRS",
    code = "EPSG:3338",
    proj4def =  "+proj=aea +lat_1=55 +lat_2=65 +lat_0=50 +lon_0=-154 +x_0=0 +y_0=0 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs",
    resolutions = 2 ^ (16:7)
)
```

transorming again to match leadlet requirements:

```{r}
pop_rgn_4326_sf <- pop_rgn_3338_sf %>% 
    st_transform(crs = 4326)
```

```{r}
m <- leaflet(options = leafletOptions(crs = epsg3338)) %>%
    addPolygons(data = pop_rgn_4326_sf,
                fillColor = "gray",
                weight = 1)

m
```



```{r}
pal <- colorNumeric(palette = "Reds", domain = pop_rgn_4326_sf$total_pop)

m <- leaflet(options = leafletOptions(crs = epsg3338)) %>%
    addPolygons(
        data = pop_rgn_4326_sf,
        fillColor = ~ pal(total_pop),
        weight = 1,
        color = "black",
        fillOpacity = 1,
        label = ~ region
    ) %>%
    addLegend(
        position = "bottomleft",
        pal = pal,
        values = range(pop_rgn_4326_sf$total_pop),
        title = "Total Population"
    )

m
```





```{r}
pal <- colorNumeric(palette = "Reds", domain = pop_rgn_4326_sf$total_pop)

m <- leaflet(options = leafletOptions(crs = epsg3338)) %>%
    addPolygons(
        data = pop_rgn_4326_sf,
        fillColor = ~ pal(total_pop),
        weight = 1,
        color = "black",
        fillOpacity = 1
    ) %>%
    addCircleMarkers(
        data = pop_4326_sf,
        lat = ~ lat,
        lng = ~ lng,
        radius = ~ log(population / 500),
        # arbitrary scaling
        fillColor = "gray",
        fillOpacity = 1,
        weight = 0.25,
        color = "black",
        label = ~ paste0(pop_4326_sf$city, ", population ", comma(pop_4326_sf$population))
    ) %>%
    addLegend(
        position = "bottomleft",
        pal = pal,
        values = range(pop_rgn_4326_sf$total_pop),
        title = "Total Population"
    )

m
```





# Conclusion

The maos show that Cook Inlet is the the highest populated region in Alaska, where Anchorage is the city with the most residents. 
 
